Using an iterated block particle filter via \pkg{spatPomp}

### 2023-04-27 ms.Rnw
minor updates, without re-running code

\title{Using an iterated block particle filter via \pkg{spatPomp}}
\author{Ning Ning and Edward L. Ionides}
\date{\today}
\begin{document}
\maketitle

The IBPF algorithm studied by \citet{ning21-ibpf} and \citet{ionides22} has been contributed to the \proglang{R} package \pkg{spatPomp} \citep{asfaw21cran,asfaw21arxiv} as the function  \code{ibpf}.
This document introduces \code{ibpf} and validates its correctness on a simple Gaussian example which is tractable using the Kalman filter.
We also test \code{ibpf} on simulated data for a measles transmission model.
In addition to the \pkg{spatPomp} code presented here, the full code to reproduce this document is available in the R Noweb (\code{.Rnw}) source file.

@@

We now proceed to carry out a similar analysis for the measles model generated by the \pkg{spatPomp} function \code{he10}.
This is a susceptible-exposed-infected-recovered model for measles transmission, described by \citet{ionides22} and \citet{asfaw21arxiv}.
For this model, exact likelihood evaluation is not available.
However, for a relatively small number of units ($U=4$) the particle filter provide an adequate approximation.
\citet{ionides22} considered fitting this model to data using IBPF, with 20 cities and up to $20\times 13$ parameters.
Here, our task is to focus on a smaller, simulated dataset, estimating fewer parameters in order to assess more clearly whether or not the block approximation is leading to substantial bias.
We choose two large towns (London and Birmingham) and two small towns (Cardiff and Hastings) since we expect that population movement from large towns to small towns is essential to explain disase persistence in small towns.
Large towns can maintain an ongoing epidemic, but, below a critical community size local extinction of the disease is expected during epidemic troughs.

>>>>

\title{Using an iterated block particle filter via \pkg{spatPomp}}
\author{Ning Ning and Edward L. Ionides}
\date{\small Compiled \today,  using \proglang{R} \Sexpr{paste0(R.version$major,".",R.version$minor)}, \pkg{spatPomp} \Sexpr{packageVersion('spatPomp')}, and \pkg{pomp} \Sexpr{packageVersion('pomp')}\\
Source code at \url{https://github.com/ionides/ibpf-tutorial}}
\begin{document}
\maketitle

The IBPF algorithm studied by \citet{ning23-ibpf} and \citet{ionides22} has been contributed to the \proglang{R} package \pkg{spatPomp} \citep{asfaw21cran,asfaw23arxiv} as the function  \code{ibpf}.
This document introduces \code{ibpf} and validates its correctness on a simple Gaussian example which is tractable using the Kalman filter.
We also test \code{ibpf} on simulated data for a measles transmission model.
In addition to the \pkg{spatPomp} code presented here, the full code to reproduce this document is available in the R Noweb (\code{.Rnw}) source file.

@@

We now proceed to carry out a similar analysis for the measles model generated by the \pkg{spatPomp} function \code{he10}.
This is a susceptible-exposed-infected-recovered model for measles transmission, described by \citet{ionides22} and \citet{asfaw23arxiv}.
For this model, exact likelihood evaluation is not available.
However, for a relatively small number of units ($U=4$) the particle filter provide an adequate approximation.
\citet{ionides22} considered fitting this model to data using IBPF, with 20 cities and up to $20\times 13$ parameters.
Here, our task is to focus on a smaller, simulated dataset, estimating fewer parameters in order to assess more clearly whether or not the block approximation is leading to substantial bias.
We choose two large towns (London and Birmingham) and two small towns (Cardiff and Hastings) since we expect that population movement from large towns to small towns is essential to explain disase persistence in small towns.
Large towns can maintain an ongoing epidemic, but, below a critical community size local extinction of the disease is expected during epidemic troughs.
